Long- and short-read metabarcoding technologies reveal similar spatio-temporal structures in fungal communities

### 2020-10-15 writing/transect_paper.Rmd
remove reviewer comments from manuscript

<!-- R1: This study seeks to determine how both sequencing target length and technology affect the spatial and temporal turnover of soil fungal communities in an African forest. The authors take an ambitious approach to these research questions by comparing multiple short-read technologies (Illumina, PacBio, and Ion Torrent) to a related long-read technology (PacBio), allowing them to make many valuable comparisons across datasets. Along with field sample replication, both spatially and temporally, the authors carefully consider a number of different bioinformatic steps for each sample, e.g. among taxonomic databases as well as whether to cluster ASV data or not. Importantly, after discovering that long read data has inherent properties that make it poorly suited for the popular DADA2 bioinformatic pipeline, the authors generate a series of new R-based tools to analyze long read data in ways that greater increase data retention. Further, they use a clever set of covariance models to improve the taxonomic assignments of ASVs that are not assigned at the bulk taxonomy step. This additional step is important in increasing the amount of sequence data that can be annotated to putative functional guilds, which in this case focuses on ECM fungi.

Overall, I found the study to be a really excellent demonstration of how different sequencing technologies and target lengths can be compared for fungi. The overall conclusion that the ecological outcome is largely similar between the different datasets is heartening. I am quite sure that longer reads will continue to increase in popularity as costs decline and this study represents an important bridge in that transition – both for its careful benchmarking but also for its presentation of tools that allow for better incorporation of long read sequence data into existing bioinformatic pipelines. I hope that my comments below help the authors make some relatively minor improvements, particular in adding more biological interpretation in a few places to help more deeply understand the patterns they have identified. -->

<!-- R2: The authors have collected an interesting data set but I feel strongly that the work requires significant improvement to the Introduction (particularly the justification) and Discussion (particularly some explanations, even speculative) sections, as detailed below in my specific comments.

Regarding the justification and conclusions, a quick scan of the literature revealed a study that used long(er) reads (covering SSU, ITS, LSU; Heeger et al. 2019 in this journal) and also separated these prior to clustering and assignment. I therefore see the contribution as stated here (line 614) as overstated except that denoising was applied prior to clustering. Phylogenetic reconstruction was also used here, although the conclusions are not justified (see my comment below). I see an important contribution of this work as developing a new method to separate fragments; however, although it should be clear (i.e. examined in the Introduction) why existing approaches like ITSx or others are not suitable.

"Long-read" has become a vague descriptor and should be clearly defined early on for this manuscript (in the abstract). How many bp or markers? I have seen the term used in metabarcoding to mean anything from a few bp longer than a MiSeq PE 2x250, to nearly 5000 bp covering multiple rRNA genes.-->

<!-- R3: This paper analyzes a transect of soils in Africa and reports how results differ from 3 differing sequencing technologies and variations to bioinformatics platforms. With sequencing technologies changing so rapidly, and so many methods to choose from, I’m sure this paper would be of interest to the Molecular Ecology Resources audience, as it provides useful data comparing sequencing technologies. I congratulate the author on herculean efforts to write so many R packages, for making his packages and scripts publicly available, and for making a pipeline that allegedly improves ASV recovery for long amplicons. I personally would find the paper more interesting if a treatment other than spatial turnover had been applied to these samples, like before and after a disturbance, or a different host tree, and one could examine if the treatment of interest gave similar inferences with different sequencing technologies. However, I believe their claim that sequencing technologies yield similar results. I do think the paper would be improved by adding in some alpha diversity analyses, which are likely to be less robust to vagaries of sequencing technologies than beta diversity analyses. I also did not see any discussion of normalization for variation in number of sequences per sample, which needs to be accounted for. Instead of figure 2 (which can be moved to the supplements), I would prefer instead to see a figure about correlations of alpha diversity among methods. I also think species accumulation curves would be useful to see. I have also suggested several ways the writing could be slightly altered to make it clearer and more engaging for the reader. The results section is quite dry, reporting the statistics rather than reporting the interesting results and using the statistics to support it. For example, “Microbial communities are spatially autocorrelated (R2 = 0.90, p <0.05)” is more interesting to read than the “Bray-curtis dissimilarity showed…” If some of the results section could be re-written to focus on the results rather than reporting statistics, it would be more engaging. I also think the introduction needs to be edited to be more cohesive with the rest of the document. For example, only abbreviations were used in the introduction but they were fully spelled out at numerous points in the methods, and this made the introduction unnecessarily difficult to follow. I also found that the discussion repeated the results a lot and thus could be truncated or re-focused to set the results more in the context of the current literature (for example some citations of similar papers was missing). Finally, I will freely admit that some portions of the methods section, like the discussion of LSUx, went over my head and were very difficult for me to understand and/or follow. I’m not sure if someone more bioinformatically inclined could do a better job reviewing some of the more technical portions or if they just need to be written more clearly. I actually spent 3x the amount of time I usually devote to reviews on this paper, and I made a lot of efforts to provide constructive criticisms of how this paper could be improved to be clearer for the reader, so I hope that they are helpful. See specific line comments below.-->

@@

ITS1 and ITS2 often have suitable variation to distinguish species, although closely related species may share identical ITS sequences in certain groups such as various Pezizomycotina [@schoch2012],
but this variablity means that they cannot be reliably aligned over the fungal kingdom [@Lindahl2013;@Tedersoo2018].
<!-- R2: This is true, but the fungi are a large and diverse group and thus it is no surprise that a highly variable rRNA marker can not be aligned. Perhaps this manuscript could discuss how meaningful a phylogenetic approach to analysis of complete fungal communities is anyway. Perhaps building a tree for each phylum is possible (and a realistic evolutionary model could be used). I think that this could be a contribution here - to explore how this variation could be dealt with using some knowledge of the clade "fungi". (Partly addressed?) -->
Additionally, the wide range of length variation of these regions may introduce bias in recovery of different taxa [@ihrmark2012;@tedersoo2015;@palmer2018].
Further bias is introduced by variation in the 5.8S region which separates the two ITS regions, as well as in the 5' end of LSU, which makes it difficult to design primers that are suitable for all fungi [@tedersoo2015].

@@

Sampling was conducted at two sites (Ang: N 9.75456° E 2.14064°; Gan: N 9.75678° E 2.31058°) approximately 30\ km apart in the _Forêt Classée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both sites were located in Soudanian savannah woodlands [@olson2001;@yorou2014] dominated by the ECM host tree _Isoberlinia doka_ (Caesalpinioideae).
At each site, 25\ soil samples were collected along a single 24 m linear transect at intervals of 1\ m in May 2015.
One third of the sample locations (3\ m spacing) were resampled one year later in June 2016, for a total of 67 samples.
For each sample, any coarse organic debris <!--Do you have any definition? Wider than 5 mm? MR-->was removed from the soil surface and a sample of approximately 5cm×5cm×5cm was extracted with a sterilized knife blade and homogenized.
A subsample of approximately 250\ mg was preserved and returned to the laboratory for extraction (Supplementary methods).

@@

<!-- R2: The Results section has many instances of the data "...are shown in Fig/Table..." This is difficult to read and I strongly suggest you summarize the findings in the text. In some cases it simply means deleting these sentences and adding the Fig/Table reference after the following sentence. It would be more helpful to talk about the overall trends here rather than listing all of the % values. -->

@@

<!-- These are mostly  results, could you summarise them here (make sure all figures are cited in the results)? This would also address the reviewers comment. -->
DADA2 denoising may perform differently on different technologies (or perhaps sequencing runs), indicated by the fact that clustering ASVs at 97% led to substantially higher correspondence between both the set of sequences recovered from the same library by different technologies and the read counts for each sequence (Figure\ \@ref(fig:read-compare)).

ASV diversity appears to be artificially inflated in the Ion Torrent dataset relative to the Illumina and PacBio datasets, which gave remarkably similar ASV richness after rarefaction, despite a difference of around 200x in unrarefied sequencing depth (Figures\ \@ref(fig:venn) and \@ref(fig:alpha-compare)).
This may be a result of the lower fraction of very high quality reads (Figure \@ref(fig:qual-check)).
We used options for DADA2 intended to improve performance on technologies with higher rates of homopolymer indel errors [@callahan2020b], but our results suggest that this still does not result in performance comparable to that which DADA2 achieves on Illumina sequences, for which it was developed [@callahan2016].

Although the longer read length capabilities of PacBio would allow recovery of longer ITS2 sequences than the other two technologies, PacBio did not recover any reads from the short amplicon library which were longer than those recovered by Illumina and Ion Torrent.
<!-- Next sentence (57 words) could be cut... -->
Notably, neither long nor short amplicon sequencing recovered any sequences identifiable to *Cantharellus*, an ECM genus which is commonly observed at the study sites as fruitbodies (personal observation), but which is also known to have accelerated evolution in the rDNA [@moncalvo2006] and longer ITS regions than other fungi [@Feibelman_1994], making it an especially difficult target for metabarcoding.
Contrary to expectations, Illumina showed a slightly higher fraction of longer ITS2 sequences than Ion Torrent, which in turn showed slightly longer sequences than PacBio (Figures\ \@ref(fig:full-length) and \@ref(fig:ITS2-length)).

@@

Assignment of ecological function to environmental fungal sequences is dependent on accurate taxonomic identification, especially at the genus level or below [@nguyen2016funguild].
However, different combinations of algortihms and reference datasets vary in their performance at confidently assigning taxonomy to sequences.
Although the RDP fungal training set and Unite performed comparably at taxonomic placement of long amplicon sequences, the Warcup database placed notably fewer sequences at all taxonomic levels for all datasets (Figure \@ref(fig:tax-chart)).
This is probably due to two factors.
First, the Warcup database does not include any non-fungi, so it cannot place any non-fungal sequences.
Second, due to its low-density coverage of the fungal kingdom (18\ 000 vs. 800\ 000 for Unite), it is likely that many ITS sequences, especially from uncultured tropical soil fungi, have no close match in the Warcup database, and so cannot be placed.
The RDP fungal training set, which has even fewer sequences (8000 fungi plus 3000 other eukaryotes), is probably more successful due to higher sequence conservation in LSU.
[@heeger2019] also found that a more conserved region, in their case 5.8S, outperformed ITS at placing sequences without close database matches.
<!-- maybe the next 2 sentences could be dropped? 41 words-->
Of the three algorithms tested, IDTAXA placed fewer sequences than RDPC or SINTAX with all databases, particularly the Warcup database, where it places <25% of ASVs even to kingdom.
This is expected given its more conservative assignment of confidence scores [@murali2018a].

>>>>

@@

ITS1 and ITS2 often have suitable variation to distinguish species, although closely related species may share identical ITS sequences in certain groups such as various Pezizomycotina [@schoch2012],
but this variablity means that they cannot be reliably aligned over the fungal kingdom [@Lindahl2013;@Tedersoo2018].
Additionally, the wide range of length variation of these regions may introduce bias in recovery of different taxa [@ihrmark2012;@tedersoo2015;@palmer2018].
Further bias is introduced by variation in the 5.8S region which separates the two ITS regions, as well as in the 5' end of LSU, which makes it difficult to design primers that are suitable for all fungi [@tedersoo2015].

@@

Sampling was conducted at two sites (Ang: N 9.75456° E 2.14064°; Gan: N 9.75678° E 2.31058°) approximately 30\ km apart in the _Forêt Classée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both sites were located in Soudanian savannah woodlands [@olson2001;@yorou2014] dominated by the ECM host tree _Isoberlinia doka_ (Caesalpinioideae).
At each site, 25\ soil samples were collected along a single 24 m linear transect at intervals of 1\ m in May 2015.
One third of the sample locations (3\ m spacing) were resampled one year later in June 2016, for a total of 67 samples.
For each sample, coarse organic debris was removed from the soil surface and a sample of approximately 5cm×5cm×5cm was extracted with a sterilized knife blade and homogenized.
A subsample of approximately 250\ mg was preserved and returned to the laboratory for extraction (Supplementary methods).

@@

@@

DADA2 denoising may perform differently on different technologies (or perhaps sequencing runs), indicated by the fact that clustering ASVs at 97% led to substantially higher correspondence between both the set of sequences recovered from the same library by different technologies and the read counts for each sequence (Figure\ \@ref(fig:read-compare)).
ASV diversity appears to be artificially inflated in the Ion Torrent dataset relative to the Illumina and PacBio datasets, which gave remarkably similar ASV richness after rarefaction, despite a difference of around 200x in unrarefied sequencing depth (Figures\ \@ref(fig:venn) and \@ref(fig:alpha-compare)).
This may be a result of the lower fraction of very high quality reads (Figure \@ref(fig:qual-check)).
We used options for DADA2 intended to improve performance on technologies with higher rates of homopolymer indel errors [@callahan2020b], but our results suggest that this still does not result in performance comparable to that which DADA2 achieves on Illumina sequences, for which it was developed [@callahan2016].

Although the longer read length capabilities of PacBio would allow recovery of longer ITS2 sequences than the other two technologies, PacBio did not recover any reads from the short amplicon library which were longer than those recovered by Illumina and Ion Torrent.
Notably, neither long nor short amplicon sequencing recovered any sequences identifiable to *Cantharellus*, an ECM genus which is commonly observed at the study sites as fruitbodies (personal observation), but which is also known to have accelerated evolution in the rDNA [@moncalvo2006] and longer ITS regions than other fungi [@Feibelman_1994], making it an especially difficult target for metabarcoding.
Contrary to expectations, Illumina showed a slightly higher fraction of longer ITS2 sequences than Ion Torrent, which in turn showed slightly longer sequences than PacBio (Figures\ \@ref(fig:full-length) and \@ref(fig:ITS2-length)).

@@

Assignment of ecological function to environmental fungal sequences is dependent on accurate taxonomic identification, especially at the genus level or below [@nguyen2016funguild].
However, different combinations of algortihms and reference datasets vary in their performance at confidently assigning taxonomy to sequences.
Although the RDP fungal training set and Unite performed comparably at taxonomic placement of long amplicon sequences, the Warcup database placed notably fewer sequences at all taxonomic levels for all datasets (Figure \@ref(fig:tax-chart)).
This is probably due to two factors.
First, the Warcup database does not include any non-fungi, so it cannot place any non-fungal sequences.
Second, due to its low-density coverage of the fungal kingdom (18\ 000 vs. 800\ 000 for Unite), it is likely that many ITS sequences, especially from uncultured tropical soil fungi, have no close match in the Warcup database, and so cannot be placed.
The RDP fungal training set, which has even fewer sequences (8000 fungi plus 3000 other eukaryotes), is probably more successful due to higher sequence conservation in LSU.
[@heeger2019] also found that a more conserved region, in their case 5.8S, outperformed ITS at placing sequences without close database matches.
Of the three algorithms tested, IDTAXA placed fewer sequences than RDPC or SINTAX with all databases, particularly the Warcup database, where it places <25% of ASVs even to kingdom.
This is expected given its more conservative assignment of confidence scores [@murali2018a].
